---
title: 'Can AUC be expressed using a case-wise criterion? No'
output: html_document
---

Consider calculating AUC for folds in which a
validation set contains $n$ observations.
To calculate AUC in the validation set, we need
the vector of prediction criteria,
$\mathbf{\hat{y}} = (\hat{y}_1, ..., \hat{y}_1)'$,
and the vector of observed responses in the validation set,
$\mathbf{y} = (y_1, ..., y_n)'$,
with 
$y_i \in \{0,1\}, \; i = 1,...,n$.

To construct the ROC curve, only the ordering of
the values in 
$\mathbf{\hat{y}}$ is needed.
Thus, assuming that there are no ties and reordering
observations if necessary, we can set 
$\mathbf{\hat{y}} = (1,2,...,n)'$.

If the AUC can be expressed as the case-wise mean or sum of a function 
$cv(\hat{y}_i,y_i)$
where $cv: \{1,2,...,n\}\times\{0,1\} \rightarrow [0,1]$, then
\begin{equation}
\label{eq:cw}
\tag{1}
\sum_{i=1}^n cv(\hat{y}_i,y_i) = AUC(\mathbf{\hat{y}},\mathbf{y})
\end{equation}
must hold for all $2^n$ possibilities for $\mathbf{y} = (y_1,...,y_n)$.
If all $y$s have the same value, either 1s or 0s, the definition
of AUC is ambiguous. It could be considered to be undefined, or
it could be set to 0 if all $y$s are 0, and to 1 if all $y$s are 1.
If AUC is considered to be undefined in these cases, we have
$2^n - 2$
admissible values for $\mathbf{y}$.

Thus, equation (\ref{eq:cw}) produces either $2^n$ or $2^n-2$ constraints.
There are, however, only $2n$ possible values for the $cv$ function.
Equation (\ref{eq:cw})
could, nevertheless, have consistent solutions.  We therefore need to verify
whether there is a value of $n$ for which (\ref{eq:cw}) has no consistent solution
for all admissible values of $\mathbf{y}$.
In that eventuality, we will have shown that AUC cannot, in general,
be expressed through a case-wise criterion.

If $n=3$, we show below that (\ref{eq:cw}) has no consistent solution
if we include all possibilities for $\mathbf{y}$, but does
if we exclude cases when all $y$s have the same value.
If $n=4$, we show that there no consistent solutions
in either case.

The following function computes AUC as function of $\mathbf{\hat{y}}$
and $\mathbf{y}$

```{r}
AUC <- function(y, yhat = seq_along(y)) {
  # AUC:
  # y is a vector of 0s and 1s
  # yhat is a predictor of y, only used for its order
  #    i.e. the result is invariant under monotonic transformations of yhat
  #    If there are ties in yhat, this function splits the ties
  #
  s <- sum(y)
  n <- length(y)
  if(s == 0) return(0)
  if(s == n) return(1)
  a <- sum(outer(yhat[y==0],yhat[y==1], `<`)) + sum(outer(yhat[y==0],yhat[y==1], `==`))/2
  a / (s*(n-s))
}
```

We generate all possible $\mathbf{y}$s of length $n$ as rows of the matrix
$\mathbf{Y}$ with the following function:

```{r}
Ymat <- function(n, exclude_identical = FALSE) {
  stopifnot(n > 0 && round(n) == n)    # n must be a positive integer
  ret <- sapply(0:(2^n - 1),
                function(x) as.integer(intToBits(x)) )[1:n, ]
  ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
  colnames(ret) <- paste0('y', 1:ncol(ret))
  if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
}
```

For $n=3$:

```{r}
Ymat(3)
```

If we exclude $\mathbf{y}$s with identical values:

```{r}
Ymat(3, TRUE)
```

The following shows $\mathbf{Y}$ with corresponding values of AUC:

```{r}
cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
```

The values of $cv(\hat{y}_i, y_i)$ that express
AUC as a sum of case-wise values are solutions of equation (\ref{eq:cw}),
which can be expressed as solutions of the following linear equation:
\begin{equation}
\label{eq:lin}
\tag{2}
(\mathbf{U} -\mathbf{Y}) \mathbf{c}_0 + \mathbf{Y} \mathbf{c}_1
=
[\mathbf{U} -\mathbf{Y} \quad \mathbf{Y}]
\begin{bmatrix}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{bmatrix}
= AUC(\mathbf{\hat{Y}},\mathbf{Y})
\end{equation}

where $\mathbf{U}$ is a matrix of 1's conformable with $\mathbf{Y}$,
$\mathbf{c}_0 = (cv(1,0), c(2,0), ..., cv(n,0))'$,
$\mathbf{c}_1 = (cv(1,1), c(2,1), ..., cv(n,1))'$,
$[\mathbf{U} -\mathbf{Y} \quad \mathbf{Y}]$ and
$\begin{bmatrix}\begin{aligned}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{aligned}
\end{bmatrix}$ are partitioned matrices, and
$\mathbf{\hat{Y}}$ is a matrix each of whose rows consists of the integers
1 to $n$.

We can test whether equation (\ref{eq:lin}) has a solution for any given $n$
by considering whether the residuals of the associated linear model,
using the design matrix
$[\mathbf{U} -\mathbf{Y} \quad \mathbf{Y}]$ to
predict the outcome
$AUC(\mathbf{\hat{Y}},\mathbf{Y})$, are zero.


```{r}
resids <- function(n, exclude_identical = FALSE, tol = 1e-8) {
  # ss
  Y <- Ymat(n, exclude_identical = exclude_identical)
  AUC <- apply(Y, 1, AUC)
  X <- cbind(1-Y, Y)
  opts <- options(warn = -1)
  on.exit(options(opts))
  fit <- lsfit(X, AUC, intercept = FALSE)
  ret <- sum(abs(fit$residuals))
  if(ret < tol){
    solution <- fit$coef
    names(solution) <- paste0('c(', c(1:n,1:n), ',', rep(0:1, each = n),')')
    attr(ret, 'solution') <- zapsmall(solution)
  }
  ret
}
```

The case $n=3$, excluding identical $y$s, has a solution:

```{r}
resids(3, exclude_identical = TRUE)
```

But, if identical $y$s are included, the equation is not consistent:

```{r}
resids(3, exclude_identical = FALSE)
```

For $n=4$, there are no solutions in either case:

```{r}
resids(4, exclude_identical = TRUE)
resids(4, exclude_identical = FALSE)
```

We have demonstated that a widely used criterion, AUC, cannot, in general, be cross-validated
using a case-wise approach.
