---
title: "Appendix: Why the complement of AUC isn't a casewise CV criterion"
output: html_document
---

Consider calculating AUC for folds in which a validation set contains $n_v$ observations. To calculate AUC in the validation set, we need the vector of prediction criteria, $\widehat{\mathbf{y}}_{(n_v \times 1)} = (\widehat{y}_1, ..., \widehat{y}_{n_v})^T$, and the vector of observed responses in the validation set, $\mathbf{y}_{(n_v \times 1)} = (y_1, \ldots, y_{n_v})^T$ with $y_i \in \{0,1\}, \; i = 1, \ldots, n_v$.

To construct the ROC curve, only the ordering of the values in 
$\mathbf{\widehat{y}}$ is relevant. Thus, assuming that there are no ties, and reordering observations if necessary, we can set 
$\mathbf{\widehat{y}} = (1, 2, \ldots, n_v)^T$.

If the AUC can be expressed as the casewise mean or sum of a function $\mathrm{cv}(\widehat{y}_i,y_i)$, where $\mathrm{cv}: \{1,2,...,n_v\}\times\{0,1\} \rightarrow [0,1]$, then
\begin{equation}
\label{eq:cw}
\tag{1}
\sum_{i=1}^{n_v} \mathrm{cv}(\widehat{y}_i,y_i) = \mathrm{AUC}(\mathbf{\widehat{y}},\mathbf{y})
\end{equation}
must hold for all $2^{n_v}$ possible values of $\mathbf{y} = (y_1,...,y_{n_v})^T$. If all $y\mathrm{s}$ have the same value, either 1 or 0, then the definition of AUC is ambiguous. AUC could be considered undefined, or it could be set to 0 if all $y$s are 0 and to 1 if all $y$s are 1. If AUC is considered to be undefined in these cases, we have $2^{n_v} - 2$ admissible values for $\mathbf{y}$.

Thus, equation (\ref{eq:cw}) produces either $2^{n_v}$ or $2^{n_v}-2$ constraints. Although there are only $2n$ possible values for the $\mathrm{cv(\cdot)}$ function, equation (\ref{eq:cw}) could, nevertheless, have consistent solutions.  We therefore need to determine whether there is a value of $n_v$ for which (\ref{eq:cw}) has no consistent solution for all admissible values of $\mathbf{y}$. In that eventuality, we will have shown that AUC cannot, in general, be expressed through a casewise sum.

If $n_v=3$, we show below that (\ref{eq:cw}) has no consistent solution if we include all possibilities for $\mathbf{y}$, but does if we exclude cases where all $y$s have the same value. If $n_v=4$, we show that there no consistent solutions in either case.

The following R function computes AUC from $\mathbf{\widehat{y}}$ and $\mathbf{y}$, accommodating the cases where $\mathbf{y}$ is all 0s or all 1s:

```{r}
AUC <- function(y, yhat = seq_along(y)) {
  s <- sum(y)
  if (s == 0) return(0)
  if (s == length(y)) return(1)
  Metrics::auc(y, yhat)
}
```

We then define a function to generate all possible $\mathbf{y}$s of length $n_v$ as rows of the matrix $\mathbf{Y}_{(2^{n_v} \times n_v)}$:

```{r}
Ymat <- function(n_v, exclude_identical = FALSE) {
  stopifnot(n_v > 0 && round(n_v) == n_v)    # n_v must be a positive integer
  ret <- sapply(0:(2^n_v - 1),
                function(x) as.integer(intToBits(x)) )[1:n_v, ]
  ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
  colnames(ret) <- paste0("y", 1:ncol(ret))
  if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
}
```

For $n_v=3$,

```{r}
Ymat(3)
```

If we exclude $\mathbf{y}$s with identical values, then

```{r}
Ymat(3, exclude_identical = TRUE)
```

Here is $\mathbf{Y}$ with corresponding values of AUC:

```{r}
cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
```

The values of $\mathrm{cv}(\widehat{y}_i, y_i)$ that express AUC as a sum of casewise values are solutions of equation (\ref{eq:cw}), which can be written as solutions of the following system of $2^{n_v}$ linear simultaneous equations in $2n_v$ unknowns:
\begin{equation}
\label{eq:lin}
\tag{2}
(\mathbf{U} -\mathbf{Y}) \mathbf{c}_0 + \mathbf{Y} \mathbf{c}_1
=
[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]
\begin{bmatrix}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{bmatrix}
= \mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})
\end{equation}
where $\mathbf{U}_{(2^{n_v} \times n_v)}$ is a matrix of 1s conformable with $\mathbf{Y}$; $\mathbf{c}_0 = [\mathrm{cv}(1,0), c(2,0), ..., \mathrm{cv}(n_v,0)]^T$; $\mathbf{c}_1 = [\mathrm{cv}(1,1), c(2,1), ..., \mathrm{cv}(n_v,1)]^T$; $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]_{(2^{n_v} \times 2n_v)}$ and
$\begin{bmatrix}\begin{aligned}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{aligned}
\end{bmatrix}_{(2n_v \times 1)}$ 
are partitioned matrices; and $\mathbf{\widehat{Y}}_{(2^{n_v} \times n_v)}$ is a matrix each of whose rows consists of the integers 1 to $n_v$.

We can test whether equation (\ref{eq:lin}) has a solution for any given $n_v$ by trying to solve it as a least-squares problem, considering whether the residuals of the associated linear model are all 0, using the "design matrix" $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ to predict the "outcome"
$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})_{(2^{n_v} \times 1)}$:

```{r}
resids <- function(n_v, exclude_identical = FALSE, 
                   tol = sqrt(.Machine$double.eps)) {
  Y <- Ymat(n_v, exclude_identical = exclude_identical)
  AUC <- apply(Y, 1, AUC)
  X <- cbind(1-Y, Y)
  opts <- options(warn = -1)
  on.exit(options(opts))
  fit <- lsfit(X, AUC, intercept = FALSE)
  ret <- max(abs(residuals(fit)))
  if(ret < tol){
    ret <- 0
    solution <- coef(fit)
    names(solution) <- paste0("c(", c(1:n_v, 1:n_v), ",", 
                              rep(0:1, each = n_v), ")")
    attr(ret, "solution") <- zapsmall(solution)
  }
  ret
}
```

The case $n_v=3$, excluding identical $y$s, has a solution:

```{r}
resids(3, exclude_identical = TRUE)
```

But, if identical $y$s are included, the equation is not consistent:

```{r}
resids(3, exclude_identical = FALSE)
```

For $n_v=4$, there are no solutions in either case:

```{r}
resids(4, exclude_identical = TRUE)
resids(4, exclude_identical = FALSE)
```

Consequently, the widely employed AUC measure of fit for binary regression cannot in general be used for a casewise cross-validation criterion.
